---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.asp = 0.818,
  out.width = "100%",
  fig.align = "center"
)
```

# mvcount

<!-- badges: start -->
[![R-CMD-check](https://github.com/rdmatheus/mvcount/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/rdmatheus/mvcount/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `mvcount` package is dedicated to estimating and analyzing goodness-of-fit within the class of regression models for count data parameterized in terms of the mean and variance parameters. The package includes implementations of some count distributions (indexed by mean and variance) such as the negative binomial, zero-inflated geometric, Bell-Touchard, BerG, and members of the class of the inflated-parameter distributions.

## Installation

You can install the current development version of `mvcount` package from [GitHub](https://github.com/rdmatheus/mvcount) with:
``` r
devtools::install_github("rdmatheus/mvcount")
```
To run the above command, it is necessary that the `devtools` package is previously installed on `R`. If not, install it using the following command:
``` r
install.packages("devtools")
```
After installing the devtools package, if you are using *Windows*, install the most current [RTools](https://cran.r-project.org/bin/windows/Rtools/) program. Finally, run the command `devtools::install_github("rdmatheus/mvcount")`, and then the package will be installed on your computer.

## Example

To illustrate the usage of the `mvcount` package, we will replicate the main features of the
application by Kokonendji et al. (2024). In the paper, Kokonendji et al. (2024) introduce the following regression models: reparameterized inflated geometric (RIGE), reparameterized inflated Bernoulli (RIBE), and reparameterized Poisson (RIPO) and compare the fits. The applicability of the proposed models is compared with the Poisson (PO), zero-inflated Poisson (ZIP), negative binomial (NB), generalized Poisson (GPO) (Consul & Jain, 1973; Rigby et al., 2019), BerG ( Bourguignon & Medeiros, 2022), and Bell-Touchard (BeTo) (Castellares et al., 2020; Lemonte, 2022) regression models, in which the mean and variance parameters index all models considered.

The data are from the area of relational psychology, initially explored by Loeys et al. (2012). They consist of a subsample of the Interdisciplinary Project for the Optimization of Separation trajectories conducted in Flanders (www.scheidingsonderzoek.be) in the context of couple separation. The data are 
available at the `COMPoissonReg` package, and can be accessed by
```{r data reading}
data("couple", package = "COMPoissonReg")
```

The dependent variable (`UPB`) is a discrete measure of the extent of unwanted pursuit behavior perpetrations displayed since the time the couple broke up for 387 individuals. Measurements are made using the Relational Pursuit-Pursuer Short Form (Cupach and Spitzberg, 2004). The explanatory variables consist of an indicator for the education level, where "1" indicates at least a bachelor's degree, and a continuous measure of anxious attachment (`anxiety`). For more details, run `help("couple", package = "COMPoissonReg")`.

```{r data_transformation_and_visualization, message=FALSE, warning=FALSE}
# Packages
library(tidyverse) # For data manipulation and visualization
library(patchwork) # Put ggplot graphs side by side
library(mvcount)   # Load the 'mvcount' package

# Some transformations:
couple <- mutate(couple, 
                 education = factor(EDUCATION, levels = c(0, 1), labels = c("No", "Yes")),
                 anxiety = ANXIETY) %>% 
  select(UPB, education, anxiety)

# Data visualization
g1 <- ggplot(couple, aes(x = UPB)) +
  theme_classic() +
  geom_bar(fill = "white", col = "black") +
  labs(x = "UPB score", y = "Frequency")

g2 <- ggplot(couple, aes(x = education, y = UPB)) +
  theme_classic() +
  geom_boxplot() +
  labs(x = "At least bachelor's degree", y = "UPB score")

g3 <- ggplot(couple, aes(x = anxiety, y = UPB, col = education)) +
  theme_classic() +
  geom_point() +
  geom_smooth(se = FALSE, col = "grey") +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "top",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7)) +
  labs(x = "Anxiety score", y = "UPB score", col = "Education")

g4 <-  mutate(couple, grp_anxiety = cut(anxiety, breaks = c(min(anxiety) - 0.5,
                                                    quantile(anxiety, c(0.25, 0.5, 0.75)),
                                                    max(anxiety) + 0.5))) %>%
  group_by(education, grp_anxiety) %>%
  summarise(mean = mean(UPB), var = var(UPB), size = n()) %>%
  ggplot(aes(x = mean, y = var, col = education, size = size)) +
  theme_classic() +
  geom_point() +
  scale_color_brewer(palette = "Set1") +
  geom_abline(intercept = 0, slope = 1, lty = 2) +
  theme(legend.position = "top",
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7)) +
  labs(x = "Sample mean", y = "Sample variance", col = "Education", size = "Group size")


(g1 + g2) / (g3 + g4)

```

We consider the following systematic components
$$
\begin{array}{ll}
\log(\mu_i)  =  \beta_0 + \beta_1\; \texttt{education}_i + \beta_2\; \texttt{anxiety}_i, &  \\
\log(\sigma_i^2) = \gamma_0 + \gamma_1\; \texttt{education}_i + \gamma_2\; \texttt{anxiety}_i, &
\end{array}
$$
where $\mu_i$ and $\sigma^2_i$ denote the mean and variance of the UPB score for the $i$th individual, $i = 1, \ldots, 387$. To fit the regression models with the `mvcount`, run the following commands:
```{r regression fits}
## For mvreg() documentation:
?mvreg

# RIGE regression
fit_rige <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "rige")

# RIBE regression
fit_ribe <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "ribe")

# RIPO regression
fit_ripo <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "ripo")

# Poisson regression
fit_po <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "po")

# ZIP regression
fit_zip <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "zip")

# NB regression
fit_nb <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "nb")

# GPO regression
fit_gpo <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "gpo")

# BerG regression
fit_berg <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "berg")

# BeTo regression
fit_beto <- mvreg(UPB ~ education + anxiety | education + anxiety, data = couple, count = "beto")
```


The goodness-of-fit measures considered in the paper for the adjusted models can be computed with:
```{r goodness-of-fit measures}

models <- c("rige", "ribe", "ripo", "po",
           "zip", "nb", "gpo", "berg", "beto")

goodness <- matrix(NA, 4, 9)
rownames(goodness) <- c("AIC", "BIC", "pseudo-R2", "Upsilon")
colnames(goodness) <- models
for(i in models){
  
    goodness[, i] <- round(c(summary(get(paste0("fit_", i)))$AIC,
                             summary(get(paste0("fit_", i)))$BIC,
                             summary(get(paste0("fit_", i)))$pseudoR2,
                             summary(get(paste0("fit_", i)))$upsilon), 4)
}

goodness

```

The normal probability plot with a simulated confidence region of the randomized quantile residuals (Dunn and Smyth, 1996) can be accessed with the `envelope()` function. Below we show the envelope graphs for each of the fits.
```{r envelope_plots, out.width="100%", fig.asp=1}
# Envelope plots
envel_rige <- envelope(fit_rige, plot = FALSE)
envel_ribe <- envelope(fit_ribe, plot = FALSE)
envel_ripo <- envelope(fit_ripo, plot = FALSE)
envel_po <- envelope(fit_po, plot = FALSE)
envel_zip <- envelope(fit_zip, plot = FALSE)
envel_nb <- envelope(fit_nb, plot = FALSE)
envel_gpo <- envelope(fit_gpo, plot = FALSE)
envel_berg <- envelope(fit_berg, plot = FALSE)
envel_beto <- envelope(fit_beto, plot = FALSE)

op <- par()$mfrow
par(mfrow = c(3, 3))
plot(envel_rige, main = "RIGE")
plot(envel_ribe, main = "RIBE")
plot(envel_ripo, main = "RIPO")
plot(envel_po, main = "PO")
plot(envel_zip, main = "ZIP")
plot(envel_nb, main = "NB")
plot(envel_gpo, main = "GPO")
plot(envel_berg, main = "BerG")
plot(envel_beto, main = "BeTO")
par(mfrow = op)
```

The final model assumes that the dependent variable has a RIPO distribution with the following systematic components:
$$
\log(\mu_i) =  \beta_0 + \beta_1 \texttt{anxiety}_i \quad \textrm{and} \quad
\log(\sigma_i^2) = \gamma_0 + \gamma_1 \texttt{education}_i + \gamma_2 \texttt{anxiety}_i,
$$
To fit the final model, run:
```{r final_fit, fig.width=6.8, fig.asp=1.5, out.width="100%"}
# Final model
fit <- mvreg(UPB ~ anxiety | education + anxiety, data = couple, count = "ripo")

# Print
fit

# Summary
summary(fit)

# Randomized quantile residuals graphical summary
layout(matrix(c(1, 3, 5, 2, 4, 5), 3, 2))
plot(fit, which = 1:5, ask = FALSE)     
layout(1)
```

Currently, the methods implemented for `"mvreg"` objects are
```{r methods}
methods(class = "mvreg")
```
## References

Bourguignon M and Medeiros RMR de (2022) A simple and useful regression model for
fitting count data. *TEST* **31**, 790–827.

Castellares F, Lemonte AJ, and Moreno–Arenas G (2020) On the two-parameter
Bell–Touchard discrete distribution. *Communications in Statistics-Theory and Methods* **49**,
4834–4852.*

Consul PC and Jain GC (1973) A generalization of the Poisson distribution. *Technometrics* **15**,
791–799.

Cupach, WR and Spitzberg, BH (2004). *The Dark Side of Relationship Pursuit: From Attraction to Obsession and Stalking*, 2nd ed. Lawrence Erlbaum Associates, Mahwah, NJ.

Dunn PK and Smyth GK (1996) Randomized quantile residuals. *Journal of Computational and Graphical Statistics* **5**, 236–244.

Kokonendji CC, Medeiros RMR and Bourguignon M (2024) Mean and variance for count regression models
based on reparameterized distributions. *Sankhya B* to appear. 

Lemonte AJ (2022) On the mean-parameterized Bell–Touchard regression model for count
data. *Applied Mathematical Modelling* **105**, 1–16.

Loeys T, Moerkerke B, De Smet O and Buysse A (2012) The analysis of zero-inflated count
data: Beyond zero-inflated Poisson regression. *British Journal of Mathematical and Statistical Psychology* **65**, 163–180.

Rigby RA, Stasinopoulos MD, Heller GZ and De Bastiani F (2019) *Distributions for Modeling Location, Scale, and Shape: Using GAMLSS in R*. CRC Press.

